---
title: "2218_Yilong_mRNA_DESeq2"
output:
  html_document:
    df_print: paged
toc: yes
---

# 1. Sample sheet

name    batch   condition       group
#################################################
s1      collA   totRNA_mock     All
s2      collA   totRNA_mock     All
s3      collB   totRNA_mock     All
#################################################
s4      collA   totRNA_arsenite treatment1
s5      collB   totRNA_arsenite treatment1
s6      collB   totRNA_arsenite treatment1
s7      collB   totRNA_HS       treatment2
s8      collB   totRNA_HS       treatment2
s9      collB   totRNA_HS       treatment2
s10     collA   totRNA_NaCl     treatment3
s11     collB   totRNA_NaCl     treatment3
s12     collB   totRNA_NaCl     treatment3
s13     collA   totRNA_UV       treatment4
s14     collA   totRNA_UV       treatment4
s15     collB   totRNA_UV       treatment4
##################################################
s16     collA   sgRNA_arsenite  treatment1
s17     collB   sgRNA_arsenite  treatment1
s18     collB   sgRNA_arsenite  treatment1
s19     collB   sgRNA_HS        treatment2
s20     collB   sgRNA_HS        treatment2
s21     collB   sgRNA_HS        treatment2
s22     collA   sgRNA_NaCl      treatment3
s23     collB   sgRNA_NaCl      treatment3
s24     collB   sgRNA_NaCl      treatment3
s25     collA   sgRNA_UV        treatment4
s26     collA   sgRNA_UV        treatment4
s27     collB   sgRNA_UV        treatment4


# 2. Load libraries and default parameters
```{r, load-libraries, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(gtools))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(grid))
#suppressPackageStartupMessages(library(apeglm))
#suppressPackageStartupMessages(library(ashr))
#suppressPackageStartupMessages(library(pheatmap))
#suppressPackageStartupMessages(library(tibble))
#library(RColorBrewer)
.pardefault <- par()
```

# 3. Load and clean the necessary data 

## 3.a) Load the data
```{r, load_data}
dfolder <- "/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/"
fcounts_gene <- "1a_mRNAseq_alignment_1SampleSheet/featureCounts/counts.tsv"

df.counts_gene <- read.delim(paste0(dfolder, fcounts_gene), header = TRUE, sep='\t', na.strings="NA", row.names = 1) 
df.counts_gene  %>% head()
```

There are `r dim(dfcounts_gene)[1]` genes/ rows and `r dim(dfcounts_gene)[2]` samples/ columns in the gene count data frame.


## 3.b) Load the metadata
```{r, label=check_counts, echo=FALSE}
fmeta_all <- "/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/data_int/sampleSheets/all_batch_corrected_sampleSheet.tsv"
df.meta_all <- read.delim(fmeta_all, header = TRUE, sep='\t', na.strings="NA", row.names = 1)
df.meta_all  %>% head()
```

## 3.c) Rearrange, clean and factor the data

There are `r dim(dfcounts_gene)[1]` genes in total.

```{r, label=cleanup}
# cleanup the metadata 
df.meta_all$condition <- gsub('arsenite', 'ars', df.meta_all$condition) 
#####################

# cleanup the data
# reorder the columns alphanumerically
df.counts_gene <- df.counts_gene[gtools::mixedorder(colnames(df.counts_gene))]

# remove the all the '_PAR_Y' rows because are empty
rows_PAR_Y <- row.names(df.counts_gene[grepl(pattern="_PAR_Y", row.names(df.counts_gene) ),])
df.counts_gene <- df.counts_gene[!row.names(df.counts_gene) %in% rows_PAR_Y,]

# remove the version number
row.names(df.counts_gene) <- gsub("\\..*","", row.names(df.counts_gene))
# are the metadata rows and data columns in the same order?
all( df.meta_all$name == colnames(df.counts_gene))

df.counts_gene  %>% head()
df.meta_all  %>% head()
```

After eliminating the PAR_Y genes, now there are still `r dim(dfcounts_gene)[1]` genes left.

```{r, factoring}
# populate the columns 'condition', 'batch' and 'group' with factors
df.meta_all[c("batch", "condition", "group")] <- lapply(df.meta_all[c("batch", "condition", "group")], factor)

#rearrange sample order:
l.sample_order <- c('totRNA_mock', 'totRNA_ars', 'totRNA_HS', 'totRNA_NaCl', 'totRNA_UV', 'sgRNA_ars', 'sgRNA_HS', 'sgRNA_NaCl', 'sgRNA_UV')
df.meta_all$condition <- factor(df.meta_all$condition, levels = l.sample_order)

df.meta_all  %>% head()
#df.meta_all$condition
```


## 4) Prepare to run DESeq2

```{r, design}
# create design
design <- ~condition + batch

# make unfiltered dds obj
dds <- DESeq2::DESeqDataSetFromMatrix(countData = df.counts_gene, colData = df.meta_all, design = design)

# transform the dds object for plotting 
rld <- rlog(dds, blind = TRUE)
```


### 4)a. PCA plots batch

#### 4)a.1. PCA plots, one at a time
```{r, plot_PCA_batch}
# add a manual color palette (9-colors)
# attach specific colors for each level of the condition column, so that for the subsets can retain the color scheme
l.colors <- c("dark grey",
              "royalblue","#00b159", "#ffc425", "red",
              "#00aedb", "darkseagreen1", "yellow", "pink")
palette.color <-setNames(l.colors, levels(rld$condition))
####################

# 1) for all samples:
rld.PCA_all <- plotPCA(rld, intgroup=c("batch", "condition"), returnData=TRUE)

percentVar <- round(100 * attr(rld.PCA_all, "percentVar"), 1)   # rounded values for each axis of PC variance

# ggplot(rld.PCA_all, aes(PC1, PC2, color=condition, shape=batch)) +
#   geom_point(aes(color=condition, fill=condition), size=4 )+
#   theme_classic(base_size = 12) +
#   scale_shape_manual(values=c(21, 24), guide=guide_legend(override.aes = list(fill="black",
#                                                             color="black", size=c(4,3))))+    # change shapes of symbols
#   xlab(paste0("PC1: ",percentVar[1])) +
#   ylab(paste0("PC2: ",percentVar[2])) +
#   # color for the edge
#   scale_colour_manual(values = c(rep("black", nlevels(rld$condition))) ) +
#   # filling color and for the legend
#   scale_fill_manual(values=palette.color,
#                     guide=guide_legend(override.aes = list(color=palette.color, shape=22, size=4) ) ) +
#   # rectangle in the plot
#   annotate("rect", xmin = 10, xmax = 45, ymin = -5, ymax = 10,
#            alpha = 0,fill = "white", color = "darkgrey", linewidth = .25, linetype = 2) +
#   #title
#   ggtitle("PCA plot for all experimental conditions" )  +
#   theme(plot.title = element_text(hjust = 0.5)) 
# ###########################

# 2) for samples with input from whole samples (totRNA)
v.cond_totRNA <- rld$condition[grepl(pattern = "totRNA_*", rld$condition), drop = TRUE]
rld.sub_totRNA <- rld[,rld$condition %in% v.cond_totRNA]
#rld.sub_totRNA$condition <- droplevels(rld.sub_totRNA$condition)

# choose colors:
sub.palette.color  <- palette.color[names(palette.color) %in% levels(v.cond_totRNA)]

rld.PCA_totRNA <- plotPCA(rld.sub_totRNA, intgroup=c("batch", "condition"), returnData=TRUE)

percentVar <- round(100 * attr(rld.PCA_totRNA, "percentVar"), 1)   # rounded values for each axis of PC variance

# ggplot(rld.PCA_totRNA, aes(PC1, PC2, color=condition, shape=batch)) +
#   geom_point(aes(color=condition, fill=condition), size=4 )+
#   theme_classic(base_size = 12) +
#   scale_shape_manual(values=c(21, 24), guide=guide_legend(override.aes = list(fill="black",
#                                                             color="black", size=c(4,3))))+    # change shapes of symbols
#   xlab(paste0("PC1: ",percentVar[1])) +
#   ylab(paste0("PC2: ",percentVar[2])) +
#   # color for the edge
#   scale_colour_manual(values = c(rep("black", nlevels(v.cond_totRNA))) ) +
#   # filling color and for the legend
#   scale_fill_manual(values=sub.palette.color,
#                     guide=guide_legend(override.aes = list(color=sub.palette.color, shape=22, size=4) ) ) +
#     #title
#   ggtitle("PCA plot for totRNA experimental conditions" )  +
#   theme(plot.title = element_text(hjust = 0.5)) 
# #############################

# 3) for samples with input from stress granules (sgRNA)
v.cond_sgRNA <- rld$condition[grepl(pattern = "sgRNA_*", rld$condition), drop=TRUE]
rld.sub_sgRNA <- rld[,rld$condition %in% v.cond_sgRNA]
#rld.sub_sgRNA$condition <- droplevels(rld.sub_sgRNA$condition)

# choose colors:
sub.palette.color  <- palette.color[names(palette.color) %in% levels(v.cond_sgRNA)]

rld.PCA_sgRNA <- DESeq2::plotPCA(rld.sub_sgRNA, intgroup=c("batch", "condition"), returnData=TRUE)

percentVar <- round(100 * attr(rld.PCA_sgRNA, "percentVar"), 1)   # rounded values for each axis of PC variance

# ggplot(rld.PCA_sgRNA, aes(PC1, PC2, color=condition, shape=batch)) +
#   geom_point(aes(color=condition, fill=condition), size=4 )+
#   theme_classic(base_size = 12) +
#   scale_shape_manual(values=c(21, 24), guide=guide_legend(override.aes = list(fill="black",
#                                                             color="black", size=c(5,3))))+    # change shapes of symbols
#   xlab(paste0("PC1: ",percentVar[1])) +
#   ylab(paste0("PC2: ",percentVar[2])) +
#   # color for the edge
#   scale_colour_manual(values = c(rep("black", nlevels(v.cond_sgRNA))) ) +
#   # filling color and for the legend
#   scale_fill_manual(values=sub.palette.color,
#                     guide=guide_legend(override.aes = list(color=sub.palette.color, shape=22, size=4) ) ) +
#   #title
#   ggtitle("PCA plot for sgRNA experimental conditions" )  +
#   theme(plot.title = element_text(hjust = 0.5)) 
```

#### 4)a.2. PCA plots, one at a time from a loop

```{r, PCA_loop_singles}
# out_dir <- "/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/2a_mRNAseq_DESeq_manual"
# 
# # par(oma = c(3,4,4,1),    # margins for whole plot
# #     mar = c(2,3,2.2,2),  # margins for individual plots
# #     mfrow = c(1, 3) )
# #####################################
# 
# l.colors <- c("dark grey",
#               "royalblue","#00b159", "#ffc425", "red",
#               "#00aedb", "darkseagreen1", "yellow", "pink")
# # set the levels as names for colors for the rld dataset, such that the other 2 being subsets, will inherit the same colors
# palette.color <-setNames(l.colors, levels(rld$condition))
# ##################
# 
# f.PCAreplot = function (rld#alpha = 0, xlab = "Lag", ylab = "Autocorrelation",
#           #grid = FALSE, ylim=c(-1.0,1.0), ...
#           ) {
#   l.intgroups <- c("batch", "condition")
#   rld$condition <- factor(rld$condition)
# 
#   # add a manual color palette (9-colors)
#   # attach specific colors for each level of the condition column, so that for the subsets can retain the color scheme
# 
#   sub.palette.color  <- palette.color[names(palette.color) %in% levels(rld$condition)]
# 
#   # run functions
#   rld.PCA <- DESeq2::plotPCA(rld, intgroup=l.intgroups, returnData=TRUE)
#   percentVar <- round(100 * attr(rld.PCA, "percentVar"), 1)   # rounded values for each axis of PC variance
# 
#   # plot the PCA
#   ggplot(rld.PCA, aes(PC1, PC2, color=condition, shape=batch)) +
#     geom_point(aes(color=condition, fill=condition), size=4 )+
#     theme_classic(base_size = 12) +
#     scale_shape_manual(values=c(21, 24), guide=guide_legend(override.aes = list(fill="black",
#                                                             color="black", size=c(4,3))))+    # change shapes of symbols
#     xlab(paste0("PC1: ",percentVar[1])) +
#     ylab(paste0("PC2: ",percentVar[2])) +
# 
#     ## keep or remove, depending if you want the initial axes or the common values of axes
#     coord_equal(xlim=c(-40, 40),ylim=c(-20,30)) +
# 
#     # color for the edge
#     scale_colour_manual(values = c(rep("black", nlevels(rld.PCA$condition))) ) +
# 
#     # filling color and for the legend
#     scale_fill_manual(values=sub.palette.color,
#                       guide=guide_legend(override.aes = list(color=sub.palette.color, shape=22, size=4) ) )
# }
# ######################################
# 
# l.namePCA2 <- list("rld", "rld.sub_sgRNA", "rld.sub_totRNA")
# l.plots <- list()
# 
# # run function and save plots to a list (1st and 2nd rows); or just show the plots (comment lines 1 &2; uncomment 3rd line)
# for (i in 1:length(l.namePCA2) ) {
#   p <- f.PCAreplot(get(l.namePCA2[[i]])) +
#     # add titles for each plot
#     ggtitle(paste0("PCA plot for ", l.namePCA2[i]) )  +
#     theme(plot.title = element_text(hjust = 0.5))
# 
#   l.plots[[i]] = p
#   #print(f.PCAreplot(get(l.namePCA2[[i]])))
# }
# 
# # save the plots one by one
# for (i in 1:3) {
#   #play with the width and height to crop some whitespace
#   ggsave(filename=paste0(l.namePCA2[i], ".png"), device ="png", path=out_dir, dpi=300, width = 8, height = 5)
#   #for more whitespace cropping:
#   #knitr::plot_crop(paste0(out_dir,'/', l.namePCA2[i], ".png"))
#   print(l.plots[[i]])
#   dev.off()
# }
```





#### 4)a.3. PCA plots, all in one image

```{r, PCA_loop_allin1}
out_dir <- "/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/2a_mRNAseq_DESeq_manual"

l.colors <- c("dark grey",
              "royalblue","#00b159", "#ffc425", "red",
              "#00aedb", "darkseagreen1", "yellow", "pink")
# set the levels as names for colors for the rld dataset, such that the other 2 being subsets, will inherit the same colors
palette.color <-setNames(l.colors, levels(rld$condition))
##################

f.PCAreplot = function (rld#alpha = 0, xlab = "Lag", ylab = "Autocorrelation",
          #grid = FALSE, ylim=c(-1.0,1.0), ...
          ) {
  l.intgroups <- c("batch", "condition")
  rld$condition <- factor(rld$condition)

  # add a manual color palette (9-colors)
  # attach specific colors for each level of the condition column, so that for the subsets can retain the color scheme

  sub.palette.color  <- palette.color[names(palette.color) %in% levels(rld$condition)]

  # run functions
  rld.PCA <- DESeq2::plotPCA(rld, intgroup=l.intgroups, returnData=TRUE)
  percentVar <- round(100 * attr(rld.PCA, "percentVar"), 1)   # rounded values for each axis of PC variance

  # plot the PCA
  ggplot(rld.PCA, aes(PC1, PC2, color=condition, shape=batch)) +
    geom_point(aes(color=condition, fill=condition), size=4 )+
    theme_classic(base_size = 12) +
    scale_shape_manual(values=c(21, 24), guide=guide_legend(override.aes = list(fill="black",
                                                            color="black", size=c(4,3))))+    # change shapes of symbols
    xlab(paste0("PC1: ",percentVar[1])) +
    ylab(paste0("PC2: ",percentVar[2])) +
  
    # for same axes:
    coord_fixed(xlim=c(-40, 40),ylim=c(-20,30)) +

    # color for the edge
    scale_colour_manual(values = c(rep("black", nlevels(rld.PCA$condition))) ) +

    # filling color and for the legend
    scale_fill_manual(values=sub.palette.color,
                      guide=guide_legend(override.aes = list(color=sub.palette.color, shape=22, size=4) ) )
}
######################################

l.namePCA2 <- list("rld", "rld.sub_sgRNA", "rld.sub_totRNA")
l.plots <- list()

# run function and save plots to a list, use lines 'p<-' and 'l.plots'
# to just show the plots (comment previous lines; uncomment 'print' line)
for (i in 1:length(l.namePCA2) ) {
  p <- f.PCAreplot(get(l.namePCA2[[i]])) + 
    # add titles for each plot
    ggtitle(paste0("PCA plot for ", l.namePCA2[i]) )  + 
    theme(plot.title = element_text(hjust = 0.5)) +
    # drop the legends
    theme(legend.position = "none") 
  
  l.plots[[i]] = p
  #print(f.PCAreplot(get(l.namePCA2[[i]])))
}

########## to arrange the legend in the middle #############

grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
  plots <- list(...)
  position <- match.arg(position)
  g <- ggplotGrob(plots[[1]] +
  #    theme(legend.position = position))$grobs
      theme(legend.position="bottom", legend.box="vertical") )$grobs
  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
  lheight <- sum(legend$height)
  lwidth <- sum(legend$width)
  gl <- lapply(plots, function(x) x +
      theme(legend.position = "none"))
  gl <- c(gl, ncol = ncol, nrow = nrow)

  combined <- switch(position,
    "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
      legend,ncol = 1,
      heights = unit.c(unit(1, "npc") - lheight, lheight)),
    "right" = arrangeGrob(do.call(arrangeGrob, gl),
      legend, ncol = 2,
      widths = unit.c(unit(1, "npc") - lwidth, lwidth)))

  grid.newpage()
  grid.draw(combined)

  # return gtable invisibly
  invisible(combined)
}

q<-grid_arrange_shared_legend(l.plots[[1]], l.plots[[2]], l.plots[[3]], ncol = 3)
ggsave(filename=paste0("overlap.png"), device ="png", path=out_dir, dpi=300, width = 16, height = 5, q) 
#add the knitr command from above to crop any remaining white space
```

### 4)b. PCA plots, no batch

```{r, PCA_loop_noBatch_allin1}
out_dir <- "/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/2a_mRNAseq_DESeq_manual"

l.colors <- c("dark grey",
              "royalblue","#00b159", "#ffc425", "red",
              "#00aedb", "darkseagreen1", "yellow", "pink")
# set the levels as names for colors for the rld dataset, such that the other 2 being subsets, will inherit the same colors
palette.color <-setNames(l.colors, levels(rld$condition))
##################

f.PCAreplot2 = function (rld#alpha = 0, xlab = "Lag", ylab = "Autocorrelation",
          #grid = FALSE, ylim=c(-1.0,1.0), ...
          ) {
  l.intgroups <- "condition"
  rld$condition <- factor(rld$condition)

  # add a manual color palette (9-colors)
  # attach specific colors for each level of the condition column, so that for the subsets can retain the color scheme

  sub.palette.color  <- palette.color[names(palette.color) %in% levels(rld$condition)]

  # run functions
  rld.PCA <- DESeq2::plotPCA(rld, intgroup=l.intgroups, returnData=TRUE)
  percentVar <- round(100 * attr(rld.PCA, "percentVar"), 1)   # rounded values for each axis of PC variance

  # plot the PCA
  ggplot(rld.PCA, aes(PC1, PC2, color=condition)) +
    geom_point(aes(color=condition, fill=condition), size=4, shape=24)+
    theme_classic(base_size = 12) +
    xlab(paste0("PC1: ",percentVar[1])) +
    ylab(paste0("PC2: ",percentVar[2])) +
  
    # for same axes:
    coord_fixed(xlim=c(-40, 40),ylim=c(-20,30)) +

    # color for the edge
    scale_colour_manual(values = c(rep("black", nlevels(rld.PCA$condition))) ) +

    # filling color and for the legend
    scale_fill_manual(values=sub.palette.color,
                      guide=guide_legend(override.aes = list(color=sub.palette.color, shape=22, size=4) ) )
}
######################################

l.namePCA2 <- list("rld", "rld.sub_sgRNA", "rld.sub_totRNA")
l.plots <- list()

# run function and save plots to a list, use lines 'p<-' and 'l.plots'
# to just show the plots (comment previous lines; uncomment 'print' line)
for (i in 1:length(l.namePCA2) ) {
  p <- f.PCAreplot2(get(l.namePCA2[[i]])) + 
    # add titles for each plot
    ggtitle(paste0("PCA plot for ", l.namePCA2[i]) )  + 
    theme(plot.title = element_text(hjust = 0.5)) +
    # drop the legends
    theme(legend.position = "none") 
  
  l.plots[[i]] = p
  #print(f.PCAreplot(get(l.namePCA2[[i]])))
}

########## to arrange the legend in the middle #############

grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
  plots <- list(...)
  position <- match.arg(position)
  g <- ggplotGrob(plots[[1]] +
  #    theme(legend.position = position))$grobs
      theme(legend.position="bottom", legend.box="vertical") )$grobs
  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
  lheight <- sum(legend$height)
  lwidth <- sum(legend$width)
  gl <- lapply(plots, function(x) x +
      theme(legend.position = "none"))
  gl <- c(gl, ncol = ncol, nrow = nrow)

  combined <- switch(position,
    "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
      legend,ncol = 1,
      heights = unit.c(unit(1, "npc") - lheight, lheight)),
    "right" = arrangeGrob(do.call(arrangeGrob, gl),
      legend, ncol = 2,
      widths = unit.c(unit(1, "npc") - lwidth, lwidth)))

  grid.newpage()
  grid.draw(combined)

  # return gtable invisibly
  invisible(combined)
}

q<-grid_arrange_shared_legend(l.plots[[1]], l.plots[[2]], l.plots[[3]], ncol = 3)
ggsave(filename=paste0("overlap_noBatch.png"), device ="png", path=out_dir, dpi=300, width = 16, height = 5, q) 
#add the knitr command from above to crop any remaining white space
#################################
```


## 5) Run DESeq and get results

```{r, run-DESeq2}
des <- DESeq(dds) 
res <- results(des)
```





```{r, MAplots}
# png("./MA_plot2.png",
#   width = 8*300,        # 5 x 300 pixels
#   height = 5*300,
#   res = 300,            # 300 pixels per inch
#   pointsize = 8)        # smaller font size
# 
# 
# .pardefault <- par()
#
# par(oma = c(3,4,4,1),    # margins for whole plot
#     mar = c(3,4,3.2,1)  # margins for individual plots
#     ) 
# 
yl <- c(-6,6)
#  
# layout(matrix(c(2, 3, 6, 4, 5, 1), 2, 3, byrow=TRUE))
# layout.show(6)
options(scipen=2) # to get scientific notation; the higher, the less the probability of lossing scientific notation

for (i in 1:length(l.objects) ) {
  DESeq2::plotMA(Rdf[[i]][[2]], ylim=yl, alpha=metadata(Rdf[[i]][[2]])$alpha/5, main=l.objects[[i]] , xlab='', ylab='', colSig='red',
                 cex.axis=1.5, cex.main=1.5)
  abline(h=c(-1,1),col="dodgerblue",lwd=2)
  }

title(main = paste0("MA plots for DE analysis, alpha=", metadata(Rdf[[i]][[2]])$alpha/5),
      xlab = "Mean expression",
      ylab = "Log fold change",
      outer = TRUE, line = 1,
      cex.lab=2, cex.main=2)

# par(.pardefault)
#dev.off()
```