---
title: "2218_Yilong_mRNA_DESeq2"
output:
  html_document:
    df_print: paged
toc: yes
---

name    batch   condition       group
#################################################
s1      collA   totRNA_mock     All
s2      collA   totRNA_mock     All
s3      collB   totRNA_mock     All
#################################################
s4      collA   totRNA_arsenite treatment1
s5      collB   totRNA_arsenite treatment1
s6      collB   totRNA_arsenite treatment1
s7      collB   totRNA_HS       treatment2
s8      collB   totRNA_HS       treatment2
s9      collB   totRNA_HS       treatment2
s10     collA   totRNA_NaCl     treatment3
s11     collB   totRNA_NaCl     treatment3
s12     collB   totRNA_NaCl     treatment3
s13     collA   totRNA_UV       treatment4
s14     collA   totRNA_UV       treatment4
s15     collB   totRNA_UV       treatment4
##################################################
s16     collA   sgRNA_arsenite  treatment1
s17     collB   sgRNA_arsenite  treatment1
s18     collB   sgRNA_arsenite  treatment1
s19     collB   sgRNA_HS        treatment2
s20     collB   sgRNA_HS        treatment2
s21     collB   sgRNA_HS        treatment2
s22     collA   sgRNA_NaCl      treatment3
s23     collB   sgRNA_NaCl      treatment3
s24     collB   sgRNA_NaCl      treatment3
s25     collA   sgRNA_UV        treatment4
s26     collA   sgRNA_UV        treatment4
s27     collB   sgRNA_UV        treatment4


## Introduction
- check if by loading all the Robject vs by running all DESeq2 from scratch, I get the same results

```{r, load_libraries}
suppressPackageStartupMessages(library(DESeq2))
```



```{r}
## TO DELETE?

setwd('/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/1a_mRNAseq_alignment_1SampleSheet')
Robj <- load(paste0(in_dir,"/DEseq_basic_DESeq.Rdata") )
ddr_shrunk
```


### Load the data 

```{r, load-data}
# example:
  # dds_expCond <- load(paste0(common_path, "DESeq2_output/DEseq_basic_DESeq.Rdata") )

# function to load all the DESeq R objects with their containing 'dds', 'ddr' and ''ddr_shr' objects into one R object
f.load_multiple_R_objects <- function(common_path, dirs, list_objs) {
  v.dt_s <- vector(mode = "list", length = length(list_objs))
  
  for (i in 1:length(list_objs) ) {
    #print(i)
    ## paste the directories and the name of the objects in the correct order!!!
    ## 'common_path' should end in a '/'!!
    load(paste0(common_path, '/', dirs[i], '/', 'DEseq_basic_DESeq.Rdata'))
    assign(paste0('dds', '_', list_objs[i]), dds) 
    assign(paste0('ddr', '_', list_objs[i]), ddr) 
    assign(paste0('ddr_shr', '_', list_objs[i]), ddr_shrunk) 
    
    v.dt_s[[i]] <- list(get(paste0('dds','_',list_objs[i])), get(paste0('ddr','_',list_objs[i])), get(paste0('ddr_shr','_',list_objs[i])) )
    names(v.dt_s)[i] <- list_objs[i]
    }
  return(v.dt_s)
}

#work_dir<- '/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/2a_mRNAseq_DESeq_manual'

# which folders to load
path_dir <- '/data/manke/group/balan/projects/Yilong_SG_Illumina_totRNAhu/output/1a_mRNAseq_alignment_1SampleSheet'
l.dir <- dir(path = path_dir, pattern = "DESeq2_", all.files = FALSE,
                          full.names = FALSE, recursive = FALSE,
                          ignore.case = FALSE, include.dirs = FALSE, no.. = FALSE)

# give names to each of the treatment
l.objects <- list('totRNA_mock_vs_ars', 'totRNA_mock_vs_HS', 'totRNA_mock_vs_NaCl', 'totRNA_mock_vs_UV')

Rdf <- f.load_multiple_R_objects(path_dir, l.dir, l.objects)
```

```{r, plot_PCA_batch}
#png("./manual_PCA.png", width=1600, height=800, units = "px", bg = "white")

rld_PCA <- plotPCA(rlog(dds_batch, blind = TRUE), intgroup=c("batch", "condition"), returnData=TRUE)
percentVar <- round(100 * attr(rld_PCA, "percentVar"), 1)   # rounded values for each axis of PC variance
ggplot(rld_PCA, aes(PC1, PC2, color=condition, shape=batch)) +
  geom_point(size=3) +
  scale_shape_manual(values=c(3, 1, 2))+    # change shapes of symbols
  scale_size_manual(values=c(8, 8, 8))+     # change sizes of symbols
  xlab(paste0("PC1: ",percentVar[1])) +
  ylab(paste0("PC2: ",percentVar[2])) 

#dev.off()
```


```{r, MAplots}
# png("./MA_plot2.png",
#   width = 8*300,        # 5 x 300 pixels
#   height = 5*300,
#   res = 300,            # 300 pixels per inch
#   pointsize = 8)        # smaller font size
# 
# par(oma = c(3,4,4,1),    # margins for whole plot
#     mar = c(3,4,3.2,1)  # margins for individual plots
#     ) 
# 
yl <- c(-6,6)
#  
# layout(matrix(c(2, 3, 6, 4, 5, 1), 2, 3, byrow=TRUE))
# layout.show(4)
options(scipen=2) # to get scientific notation; the higher, the less the probability of lossing scientific notation

for (i in 1:length(l.objects) ) {
  DESeq2::plotMA(Rdf[[i]][[2]], ylim=yl, alpha=metadata(Rdf[[i]][[2]])$alpha/5, main=l.objects[[i]] , xlab='', ylab='', colSig='red',
                 cex.axis=1.5, cex.main=1.5)
  abline(h=c(-1,1),col="dodgerblue",lwd=2)
  }

title(main = paste0("MA plots for DE analysis, alpha=", metadata(Rdf[[i]][[2]])$alpha/5),
      xlab = "Mean expression",
      ylab = "Log fold change",
      outer = TRUE, line = 1,
      cex.lab=2, cex.main=2)

#dev.off()
```
